Overlapping Batches for the Assessment of Solution Quality in Stochastic Programs

### 2013-10-10 omrp.tex
Increased font size to 11 pt
Updated Guzin's phone number
Add test results from DB1
	new file:   images/db1_varvar.png
	modified:   omrp.bib
	modified:   omrp.tex

\documentclass{article}

@@

% AUTHOR: Enter the authors of the article, see end of the example document for further examples
\author{David~Love$^{a}$\ \ and G\"{u}zin~Bayraksan$^{b}$\thanks{Corresponding author. Tel.: +1 520-621-2605. Fax:+1 614-292-7852}\\[6pt]
{\small
      $^{a}$ Program in Applied Mathematics, University of Arizona, Tucson, AZ 85721, USA, \texttt{dlove@math.arizona.edu}} \\
{\small 
      $^{b}$ Integrated Systems Engineering, The Ohio State University, Columbus, OH 43210, USA, \texttt{bayraksan.1@osu.edu}}}

@@

The rest of the paper is organized as follows.  
In the next section, we review relevant background information.  
We first discuss the overlapping batch means in \S \ref{ssec:obm} and then briefly go over the multiple replications procedure in \S \ref{ssec:mrp}.  
In \S \ref{sec:omrp}, we present the overlapping multiple replications procedure (OMRP) with variable overlap.  
In \S \ref{sec:theory}, we prove several asymptotic properties of the OMRP estimators and in \S \ref{sec:comp}, we test the performance of OMRP on three problems.  
We end in \S \ref{sec:concl} with a summary and conclusions.

@@

We empirically tested the effectiveness of OMRP on three problems. 
The first test problem is a newsvendor problem with demand distributed uniformly on $\{0, 0.0001, 0.0002, \dots, 10\}$, the second is a capacity expansion planning problem, denoted CEP1, with random demand vector $\xit$ of dimension 3, with 216 total realizations, and the final is a stochastic knapsack problem denoted 10D, with 10 stochastic parameters and 10 binary decision variables.  
A description of CEP1 can be found in \cite{higle1996}, and a description of 10D is available in \cite{kleywegt2002sample}. 
The instance of the newsvendor problem as well as the candidate solutions used for both problems can be found in \cite{Bayraksan2006}, except here we used a discrete uniform demand distribution for the newsvendor problem.  
Both CEP1 and newsvendor can be written as two-stage stochastic linear programs with recourse. 
CEP1 was solved with the regularized decomposition algorithm of \cite{ruszczynski1986regularized}, using the accelerated version of the algorithm implemented in C++ \cite{ruszczynski1997accelerating}. 
The newsvendor problem was solved using the quicksort algorithm in C++, and 10D was solved using dynamic programming.  
In all three cases, we used the Mersenne Twister algorithm to generate random samples \cite{wagner_twister_09}.

We used a total sample size of $n = 30 m$, with several values of the batch size $m$. 
Then, at each value of $m$, we formed OMRP estimators, with varying amounts of overlap. 
We also formed 90\% confidence intervals on the optimality gap using ORMP estimators. 
Results for the newsvendor problem were compiled over 10,000 independent runs, while the CEP1 and 10D results were taken over 1,000 independent runs.  
Summary of results from these experiments are depicted in Figures~\ref{fig:nv} and \ref{fig:varvar}.

@@

\begin{figure}[htb!]
	\centering
	\subfloat[][CEP1]{
	   \label{fig:cep1_varvar}
	   \includegraphics[width=.32\textwidth]{images/cep1_varvar}}
	\subfloat[][10D]{
	   \label{fig:10d_varvar}
	   \includegraphics[width=.32\textwidth]{images/10D_varvar}}
		\caption{
		Variance Reduction for
		(a) CEP1 and 
		(b) 10D
		for various values of $\gamma / m$.
		}
\label{fig:varvar}
\end{figure}

Figures~\ref{fig:nv}(a) and \ref{fig:varvar} show the reduction of variance, with each term being normalized with respect to $\var{nVG}$ of the nonoverlapping case (MRP). 
The solid lines in these figures show the theoretical variance reduction from \cite{Welch1987} given by (\ref{eq:var_reduct_formula}). 
Empirical results agree with the theoretical values well, with more variability observed in CEP1 and 10D, providing evidence that similar variance reduction can be achieved in stochastic programming by overlapping the batches even with small sample sizes.

@@

Finally, Figure~\ref{fig:nv}(c) shows the coverage probability of confidence intervals generated by OMRP for several values of $m$ across varying values of $\gamma/m$ for the newsvendor problem.  
The results of the classical overlapping batches estimators show that coverage probability does not change with the amount of overlap, which we can  empirically see in this figure for OMRP for newsvendor, and we observed the same for CEP1 and 10D.  
The coverage probabilities from applying the (nonoverlapping) MRP algorithm to these problems presented in \cite{Bayraksan2006} agree with the our results: for the newsvendor problem, coverage probability drops as $m$ increases and the coverage probability of CEP1 (not shown) remains fairly constant around the desired value of $90\%$. 
Coverage probability for 10D (not shown) was very high, more than $0.99$.  The bias from solving (\ref{eq:sto_prog_m}) for 10D seems to be much more significant than the variance.

>>>>

\documentclass[11pt]{article}

@@

% AUTHOR: Enter the authors of the article, see end of the example document for further examples
\author{David~Love$^{a}$\ \ and G\"{u}zin~Bayraksan$^{b}$\thanks{Corresponding author. Tel.: +1 614-292-1695. Fax:+1 614-292-7852}\\[6pt]
{\small
      $^{a}$ Program in Applied Mathematics, University of Arizona, Tucson, AZ 85721, USA, \texttt{dlove@math.arizona.edu}} \\
{\small 
      $^{b}$ Integrated Systems Engineering, The Ohio State University, Columbus, OH 43210, USA, \texttt{bayraksan.1@osu.edu}}}

@@

The rest of the paper is organized as follows.  
In the next section, we review relevant background information.  
We first discuss the overlapping batch means in \S \ref{ssec:obm} and then briefly go over the multiple replications procedure in \S \ref{ssec:mrp}.  
In \S \ref{sec:omrp}, we present the overlapping multiple replications procedure (OMRP) with variable overlap.  
In \S \ref{sec:theory}, we prove several asymptotic properties of the OMRP estimators and in \S \ref{sec:comp}, we test the performance of OMRP on four problems.  
We end in \S \ref{sec:concl} with a summary and conclusions.

@@

We empirically tested the effectiveness of OMRP on four problems.
The first test problem is a newsvendor problem with demand distributed uniformly on $\{0, 0.0001, 0.0002, \dots, 10\}$, the second is a capacity expansion planning problem, denoted CEP1, with random demand vector $\xit$ of dimension 3, with 216 total realizations, the third is a stochastic knapsack problem denoted 10D, with 10 stochastic parameters and 10 binary decision variables, and the final is a stochastic vehicle routing problem denoted DB1, with 46 stochastic parameters and a total of $4.5 \cdot 10^{25}$ realizations.
Descriptions of CEP1, 10D, and DB1 can be found in \cite{higle1996}, \cite{kleywegt2002sample}, and \cite{donohue1995upper}, respectively.
The instance of the newsvendor problem as well as the candidate solutions used for both problems can be found in \cite{Bayraksan2006}, except here we used a discrete uniform demand distribution for the newsvendor problem.  
The problems CEP1, DB1 and newsvendor can be written as two-stage stochastic linear programs with recourse.
CEP1 and DB1 were solved with the regularized decomposition algorithm of \cite{ruszczynski1986regularized}, using the accelerated version of the algorithm implemented in C++ \cite{ruszczynski1997accelerating}. 
The newsvendor problem was solved using the quicksort algorithm in C++, and 10D was solved using dynamic programming.  
In all cases, we used the Mersenne Twister algorithm to generate random samples \cite{wagner_twister_09}.

We used a total sample size of $n = 30 m$, with several values of the batch size $m$. 
Then, at each value of $m$, we formed OMRP estimators, with varying amounts of overlap. 
We also formed 90\% confidence intervals on the optimality gap using ORMP estimators. 
Results for the newsvendor problem were compiled over 10,000 independent runs, while the CEP1 and 10D results were taken over 1,000 independent runs, and DB1 results were generated over 500 independent runs.
Summary of results from these experiments are depicted in Figures~\ref{fig:nv} and \ref{fig:varvar}.

@@

\begin{figure}[htb!]
	\centering
	\subfloat[][CEP1]{
	   \label{fig:cep1_varvar}
	   \includegraphics[width=.32\textwidth]{images/cep1_varvar}}
	\subfloat[][10D]{
	   \label{fig:10d_varvar}
	   \includegraphics[width=.32\textwidth]{images/10D_varvar}}
	\subfloat[][DB1]{
	   \label{fig:db1_varvar}
	   \includegraphics[width=.32\textwidth]{images/db1_varvar}}
		\caption{
		Variance Reduction for
		(a) CEP1 and 
		(b) 10D
		(c) DB1
		for various values of $\gamma / m$.
		}
\label{fig:varvar}
\end{figure}

Figures~\ref{fig:nv}(a) and \ref{fig:varvar} show the reduction of variance, with each term being normalized with respect to $\var{nVG}$ of the nonoverlapping case (MRP). 
The solid lines in these figures show the theoretical variance reduction from \cite{Welch1987} given by (\ref{eq:var_reduct_formula}). 
Empirical results agree with the theoretical values well, with more variability observed in CEP1 and 10D, providing evidence that similar variance reduction can be achieved in stochastic programming by overlapping the batches even with small sample sizes.
DB1 shows some variance reduction, but less than the asymptotic amount for small sample sizes.

@@

Finally, Figure~\ref{fig:nv}(c) shows the coverage probability of confidence intervals generated by OMRP for several values of $m$ across varying values of $\gamma/m$ for the newsvendor problem.  
The results of the classical overlapping batches estimators show that coverage probability does not change with the amount of overlap, which we can  empirically see in this figure for OMRP for newsvendor, and we observed the same for CEP1 and 10D.  
The coverage probabilities from applying the (nonoverlapping) MRP algorithm to these problems presented in \cite{Bayraksan2006} agree with the our results: for the newsvendor problem, coverage probability drops as $m$ increases and the coverage probability of CEP1 (not shown) remains fairly constant around the desired value of $90\%$. 
Coverage probability for 10D and DB1 (not shown) were very high, more than $0.99$.  The bias from solving (\ref{eq:sto_prog_m}) for these larger problems seems to be much more significant than the variance.
